---
title: "Instacart Market Basket Analysis"
author: "Sam Playle"
date: "28 June 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This analysis is based on Instacart's dataset of 3 million anonymized orders, which is discussed in a blog post here: https://tech.instacart.com/3-million-instacart-orders-open-sourced-d40d29ead6f2

The goal of the Kaggle competition is to predict which items, from among those previously purchased, will be included in a customer's next order. 
https://www.kaggle.com/c/instacart-market-basket-analysis



## Exploratory analysis

### Data

The data comprise five .csv files.
```{r echo = T}
data <- list.files("../Data")
data
```


Let's have a look at them. The first (alphabetically) is `aisles.csv`.
```{r echo = T}
aisles <- read.csv("../Data/aisles.csv")
dim(aisles)
head(aisles)
```

We see there are 134 "aisles", each of which corresponds to a different collection of groceries. 

The next .csv file is `departments.csv`.
```{r echo = T}
depts <- read.csv("../Data/departments.csv")
dim(depts)
head(depts)
```

So there are 21 departments.

Next we have the biggest .csv files. `order_products__prior.csv` contains previous order contents for all customers. 


```{r echo = T}
opp <- read.csv(file = "../Data/order_products__prior.csv")
dim(opp)
head(opp)
```


```{r echo = T}
opt <- read.csv(file = "../Data/order_products__train.csv")
dim(opt)
head(opt)
```


```{r echo = T}
orders <- read.csv(file = "../Data/orders.csv")
dim(orders)
head(orders)
```

```{r echo = T}
products <- read.csv(file = "../Data/products.csv")
dim(products)
head(products)
```


